---
title: "Pricing of Most Abused & Most Popular Drugs"
---
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidytext)

library(broom)
library(dplyr)
library(forcats)
library(httr)
library(janitor)
library(knitr)
library(plotly)
library(readxl)
library(rvest)
library(stringr)
library(viridis)
final_dataset = read_csv("./data/final_dataset_20171205.csv")
#plotly formatting
f = list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f"
)
trade_name_axis = list(
  title = "Trade Name",
  titlefont = f
)
avg_price_axis = list(
  title = "Average Drug Price",
  titlefont = f
)
most_popular_drugs = c("Viagra", "Lipitor", "Xanax", "Celebrex", "Prozac", "Crestor", "Nexium", "Lisinopril", "Zoloft", "Abilify", "Prilosec", "Percocet", "Vicodin", "Zocor")
most_abused_drugs = c("Vicodin", "OxyContin", "Ritalin", "Concerta", "Focalin", "Metadate", "Ambien", "Lunesta", "Valium", "Xanax", "Duragesic", "Opana", "Darvon", "Dilaudid", "Demerol", "Lomotil", "Nembutal", "Dexedrine", "Adderall", "Percocet", "Suboxone", "Fentanyl") 
```

#
#

##### The Distribution of Average Price for Most Abused Drugs 

  + Ambien and Lunesta, which both treat sleep disorders, have the highest median of average drug price of $13.45 and $11.94 respectively. Focalin and Ritalin, which both treat ADHD, have the lowest median of average drug price of $0.97 each. 

```{r Case Study Analysis, warning = FALSE, message = FALSE}
final_dataset %>%
  filter(!is.na(avg_price)) %>%
  filter(trade_name %in% most_abused_drugs) %>%
  group_by(trade_name) %>% 
  plot_ly(y = ~avg_price, type = "box", color = ~trade_name, colors = "Set3") %>%
  layout(title = "Distribution of Average Price for Most Abused Drugs", xaxis = trade_name_axis, yaxis = avg_price_axis)
```

#
#

##### The Distribution of Average Price for Most Popular Drugs

  + Viagra, which treats erectile dysfunction (acute), had the highest median of average price of $35.47. 
  
  + Xanax, which treats anxiety and panic attack (acute), had the lowest median of average price of $3.41.
  
  + There is no clear trend between disease type (acute vs. chronic) and disease condition (i.e: psychiatric drugs vs. cardiovascular).
  
```{r Most Popular, warning = FALSE, message = FALSE}
final_dataset %>%
  filter(!is.na(avg_price)) %>%
  filter(trade_name %in% most_popular_drugs) %>%
  plot_ly(y = ~avg_price, type = "box", color = ~trade_name, colors = "Set3") %>%
  layout(title = "Distribution of Average Price for Popular Drugs", xaxis = trade_name_axis, yaxis = avg_price_axis)
```
