---
title: "Most Advertised Drugs & Pricing"
output: 
  html_document:
    code_folding: hide
---


```{r load_packages, echo = FALSE, results = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidytext)

library(broom)
library(dplyr)
library(forcats)
library(httr)
library(janitor)
library(knitr)
library(magrittr)
library(plotly)
library(readxl)
library(rvest)
library(stringr)
library(viridis)
```

```{r load_data, echo = FALSE, results = FALSE, warning = FALSE, message = FALSE}
final_dataset = read_csv("./data/final_dataset_20171205.csv") 
final_NIH_funding = read_csv("./data/final_NIH_funding.csv") 
drug_price = read_csv("../data/drug_price.csv") 

#plotly formatting
f = list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f"
)
trade_name_axis = list(
  title = "Trade Name",
  titlefont = f
)
avg_price_axis = list(
  title = "Average Drug Price",
  titlefont = f
)
top_advertising_cost = c("Humira", "Lyrica", "Eliquis", "Xeljanz", "Cialis", "Invokana", "Breo", "Latuda", "Victoza", "Viagra")
```

##### The Distribution of Average Price for Most Advertised Drugs, Ordered by Descending Advertising Spending

  + Elliquis had the highest advertising spending of $174 million, yet had the lowest median of average drug price of $5.28. 
  
  + Xeljanz had the highest median of average drug price of $43.14, among this group. Also, Xeljanz had the most variability in this group.
  
```{r Most Advertised plots, warning = FALSE, message = FALSE}
#shiny application : most advertised (by spending) vs. most abused vs. most prescribed (by sales)
#another shiny application: give list of all "most" drugs and can see boxplot for that specific drug
final_dataset %>%
  filter(!is.na(avg_price)) %>%
  filter(trade_name %in% top_advertising_cost) %>%
  ungroup() %>%
  mutate(trade_name = factor(trade_name)) %>%
  mutate(trade_name_new = fct_relevel(trade_name, c("Eliquis", "Xeljanz", "Cialis", "Invokana", "Latuda","Viagra"))) %>%
   group_by(trade_name) %>% 
  plot_ly(x = ~trade_name_new, y = ~avg_price, type = "box", color = ~trade_name, colors = "Set3") %>%
  layout(title = "Distribution of Average Price for Most Advertised Drugs, 
         Ordered by Descending Advertising Spending", xaxis = trade_name_axis, yaxis = avg_price_axis)
```
**We see a weak trend: The less advertising, the higher the price**