---
title: "What Drives Drug Prices?"
author: "MeOak/Muhire/Manali/Josie"
date: "12/4/2017"
output: html_document
---
### Introduction

In 2016, 1 in 5 Americans reported taking four or more prescription drugs (“Public Opinion on Prescription…”, 2016). Also, high drug prices have become an increasing concern in the United States (Hancock, 2016 ; "Why Prescription Drugs...”, 2017). This project aims at  exploring some of the factors associated with US prescription drug prices including  disease research funding, drug need (prevalence), sales, and advertising. Additionally, we seek to understand the types of drugs that receive patents/exclusivity.

### The Data

* FDA Orange Book: This contained 4 datasets: patents, exclusivity, ingredients, and products.

  + Patents: Contains patent number and expiration dates of FDA approved drugs.
  
  + Exclusivity: Contains information on drugs that were granted exclusivity by the FDA.  Exclusivity provides protection from new market competition for a short time.
  
  + Ingredients: Contained information on the active ingredients in drugs.
  
  + Products: Contained the trade name for the drug, the company that developed the drug, and whether the drug was brand or generic.
  
* Drug Price: This dataset contained the prices paid by pharmacies to acquire the drug.

* NIH Funding: Contained the NIH finding for disease research areas for 2013 to 2018 (estimated) and the prevalence of the disease.  

* Drug Disease: Dataset linking drugs was the disease that they treat.

* Advertising: Money spent on advertising on TV in 2016 for the top 10 most advertised drugs.

* Sales: Top 20 prescription drugs based on US sales in 2016 (in billions of dollars)

### The Variables
The variables that we retained are: 

* trade_name: the trade name of the drug

* drug_type: the drug type (B: Brand or G: Generic)

* otc: over the counter drug (Yes or No)

* pricing_unit: EA for each, GM for gram, and ML for milliliters

* year: the funding and pricing year

* avg_price: the average price for the particular year

* dose_form: the dose form (tablet, solution, solution injectable)

* route: route of administration (oral or IV)

* strength: the strength of the dose

* application_no: the FDA assigned number to the drug application.

* product_no: The FDA assigned number to identify the application products; used for joining datasets.

* approval_year: the year the patent was approved

* app_full: the full name of the company that developed the drug

* active_ingredient: the active ingredient

* ingredient: the active ingredients in a wide format; used for joining datasets

* patent_num: the patent number

* pediatric_exclusivity_granted: yes or no

* year_pat_expires: the patent expiration year

* year_excl_expires: the exclusivity expiration year

* disease: the disease the drug treats

* prevalence: the disease prevalence for 2015

* US_mortality_2015: the disease mortality rate

* funding: the NIH funding for disease research in millions

* disease_type: the type disease (acute or chronic)

### Data Wrangling 
* The code chunks below were used for loading and tidying the data.

  + We put all sub-datasets in long format. We renamed variables as necessary. 
 
  + We restricted our final dataset to contain only non-solution drugs. These drugs are described as “each” or “per gram.” We omitted all solution-type drugs because the Drug Price dataset did not specify if liquid drugs were drinkable solutions or injectable.  Also all drugs were measured by two different units: mills and grams. Therefore, we could not adequately compare between solution and per gram/each type of drugs. 
 
  + The Drug Price Excel dataset had over 5 million observations. To representatively condense this dataset, we averaged monthly drug price for each drug into a yearly “avg_price” variable.  Essentially, each unique drug observation had 12 drug prices for each month of a year. We averaged these 12 drug prices into one average price for the year. This condensed the Drug Price dataset to approximately 2.5 million observations. 
 
  + We joined patents, products, exclusivity, ingredient, drug_disease_data, final_nih_funding, and drug_price and chronic_diseases_final into our Final Dataset named final_dataset. We used the variables ingredient, application_no, product_no, drug_type, trade_name, otc, and disease as keys for joining (inner and left).

```{r load_packages, message = FALSE}
library(tidyverse)
library(dplyr)
library(rvest)
library(httr)
library(janitor)
library(stringr)
library(forcats)
library(tidytext)
library(viridis)
library(broom)
library(plotly)
library(forcats)
library(readxl)
```

#### **Dataset 1: Drug Prices**

* Scrapping and Saving as CSV the Drug Price to speed up run-time.
```{r tidy_drug_price}
# # API code
# url = "https://data.medicaid.gov/resource/tau9-gfwr.csv"
# drug_price =
#   GET(url, query = list("$limit" = 9999999)) %>%
#   content("parsed") %>%
#   clean_names()

# # CSV code
drug_price =
  read_csv("../data/drug_price.csv") %>%
  clean_names()
```

* The raw `drug_price` dataset is huge and takes a while to load.
* It contains weekly drug prices from november 2013 to october 2017 recorded. There are `r drug_price %>% nrow()` observations and `r drug_price %>% ncol()` variables in the dataset:
    + `as_of_date`: date of data posting
    + `classification_for_rate_setting`: indicates whether, for NADAC calculation, the NDC was considered brand (‘B’) or generic (‘G’) or brand (‘B’) approved under an Abbreviated New Drug Application (ANDA) (‘B-ANDA’).
    + `corresponding_generic_drug_effecitve_date`:  effective date of when the Corresponding Generic Drug NADAC Per Unit is assigned to a multiple source brand drug NDC
    + `corresponding_generic_drug_nadac_per_unit`: NADAC for the corresponding generic drug
    + `effective_date`: effective date of the NADAC Per Unit cost
    + `explanation_code`: codes that pertain to how the NADAC was calculated. see explanation code
descriptions [here](https://www.medicaid.gov/medicaid-chip-program-information/by-topics/prescription-drugs/ful-nadac-downloads/nadacdatadefinitions.pdf)
    + `nadac_per_unit`: National Average Drug Acquisition Cost per unit
    + `ndc`:  11-digit FDA code that includes the labeler code, product code, and package code
    + `ndc_description`: drug name, strength, and dosage form of the drug 
    + `otc`: over-the-counter (OTC) product (‘Y’ or ‘N’)
    + `pharmacy_type_indicator`: source of pharmacy survey data used to calculate the NADAC. ‘ C/I’ indicates data was collected from surveys of Chain/Independent pharmacies. Other pharmacy type indicators are not used at this time.
    + `pricing_unit`:  pricing unit for the associated NDC (‘ML’, ‘GM’ or ‘EA’)
* We will focus on the variables `effective_date`, `ndc`, `ndc_description`, `otc`, `classification_for_rate_setting`, `pricing_unit`, and `nadac_per_unit`. We will rename the variables `effective_date`, `ndc_description`, `classification_for_rate_setting`, and `nadac_per_unit` as `date`, `drug_name`, `drug_type`, and `unit_price` respectively. We will also split the `date` variable into the `year`, `month` and `date` variables.

* Cleaning Drug Price Dataset
```{r rename_split_date}
drug_price = 
  drug_price %>%
  rename(trade_name = ndc_description,
         drug_type = classification_for_rate_setting,
         unit_price = nadac_per_unit) %>%
  mutate(trade_name = str_replace(trade_name, "[0-9]", "_"),
         trade_name = str_to_title(trade_name),
         otc = recode(otc, "Y" = "yes", "N" = "no"),
         otc = str_to_title(otc),
         otc = as.factor(otc)) %>% 
  separate(trade_name, into = c("trade_name", "remove"), sep = "_") %>% 
  separate(effective_date, into = c("year", "month", "date")) %>%
  select(-c(date, x1, remove)) %>%
  mutate(trade_name = trimws(trade_name, which = c("both"))) %>% 
  separate(trade_name, into = c("trade_name", "remove"), sep = " ") %>% 
  filter(is.na(remove)) %>%
  select(-remove) %>% 
  filter(pricing_unit == "GM" | pricing_unit == "EA") %>% 
  select(trade_name, everything()) 
```

```{r}
new_drug_price = drug_price %>% 
  group_by(trade_name, drug_type, otc, pricing_unit, year) %>% 
  summarize(avg_price = mean(unit_price, na.rm = TRUE))

new_drug_price
```

#### **Dataset 2: Funding for Research**

* The NIH RCDC Report
* This dataset contains the annual support for various disease categories based on grants, contracts, and other funding mechanisms used across the National Institutes of Health (NIH).  Prevalence and mortality data included in this dataset is from the National Center for Health Statistics (NCHS) at the Centers for Disease Control & Prevention (CDC).
*Funding for 2018 are estimated figures.
```{r load_funding}
url = "https://report.nih.gov/categorical_spending.aspx"
NIH_xml = read_html(url)
NIH_funding = (NIH_xml %>% html_nodes(css = "table"))[[3]] %>% html_table() %>% as_tibble()
NIH_funding = NIH_funding %>%
  clean_names() %>%
  mutate(US_mortality_2015 = x2015_us_mortality_19)

NIH_funding$x2015_us_prevalence_standard_error_19  = recode(NIH_funding$x2015_us_prevalence_standard_error_19 , "-" = "NA")
NIH_funding$US_mortality_2015 = recode(NIH_funding$US_mortality_2015, "-" = "NA")

NIH_funding = NIH_funding %>%
  separate(x2015_us_prevalence_standard_error_19, into = c("prevalence", "remove"), sep = "\\%") %>%
  select(-x2015_us_mortality_19, -remove)
NIH_funding$fy_2013actual = recode(NIH_funding$fy_2013actual, "+" = "NNA")
NIH_funding$fy_2014actual = recode(NIH_funding$fy_2014actual, "+" = "NNA")
NIH_funding$fy_2015actual = recode(NIH_funding$fy_2015actual, "+" = "NNA")
NIH_funding$fy_2016actual = recode(NIH_funding$fy_2016actual, "+" = "NNA")

NIH_funding = NIH_funding %>%
  mutate(fy_2013 = as.numeric(substring(fy_2013actual, 2))) %>%
  mutate(fy_2014 = as.numeric(substring(fy_2014actual, 2))) %>%
  mutate(fy_2015 = as.numeric(substring(fy_2015actual, 2))) %>%
  mutate(fy_2016 = as.numeric(substring(fy_2016actual, 2))) %>%
  mutate(fy_2017 = as.numeric(substring(fy_2017estimated_enacted, 2))) %>%
  mutate(fy_2018estimated = as.numeric(substring(fy_2018estimated, 2))) 

NIH_funding$prevalence = as.numeric(NIH_funding$prevalence)
NIH_funding$US_mortality_2015 = as.numeric(NIH_funding$US_mortality_2015)
  
NIH_funding = NIH_funding %>%
  select(-c(fy_2013actual, fy_2014actual, fy_2015actual, fy_2016actual, fy_2017estimated_enacted))

final_NIH_funding = NIH_funding %>%
  rename(disease = research_disease_areas_dollars_in_millions_and_rounded, 
         '2013' = fy_2013,
         '2014' = fy_2014,
         '2015' = fy_2015,
         '2016' = fy_2016,
         '2017' = fy_2017,
         '2018' = fy_2018estimated) %>%
  gather(key = "year", value = funding, c("2013", "2014", "2015", "2016", "2017", "2018")) %>% 
  mutate(disease = str_replace(disease, "[0-9]", "_"),
         disease = str_replace(disease, "\\-", ""),
         disease = str_replace(disease, "ALS", "Amyotrophic Lateral Sclerosis"),
         disease = str_replace(disease, "Sleep Research", "Sleep Disorders"),
         disease = str_replace(disease, "Smoking and Health", "Smoking Cessation")) %>% 
  separate(disease, into = c("disease", "remove"), sep = "\\_") %>% 
  select(-remove) %>% 
  separate(disease, into = c("disease", "remove"), sep = "\\(") %>% 
  select(-remove)

rm(NIH_funding)

final_NIH_funding
```

#### **Dataset 3: Patents**

* The Patents dataset from the FDA Orange Book
* Below is a description of the dataset and variables from https://www.fda.gov/Drugs/InformationOnDrugs/ucm129689.htm

* New Drug Application Type: The type of new drug application approval, either New Drug Applications(NDA or innovator) or Abbreviated New Drug Applications(ANDA or generic)

* New Drug Application (NDA) Number (appl_no) : The FDA assigned number to the application. 

* Product Number: The FDA assigned number to identify the application products.  Each strength is a separate product.  May repeat for multiple part products. 

* Patent Number (patent_num): Patent numbers as submitted by the applicant holder for patents covered by the statutory provisions.  May repeat for multiple applications and multiple products. 

* Pediatric Exclusivity Granted: Yes or No.  Was pediatric exclusivity granted by the agency?

* Patent Expire Date: The date the patent expires as submitted by the applicant holder including applicable extensions.  This was split into Year, Month, and Day variables

* Drug Substance Flag: Patents submitted on FDA Form 3542 and listed after August 18, 2003 may have a drug substance flag indicating the sponsor submitted the patent as claiming the drug substance.   

* Drug Product Flag: Patents submitted on FDA Form 3542 and listed after August 18, 2003 may have a drug product flag indicating the sponsor submitted the patent as claiming the drug product.   

* Patent Use Code: Code to designate a use patent that covers the approved indication or use of a drug product.  May repeat for multiple applications, multiple products and multiple patents.

* Patent Delist Request Flag: Sponsor has requested patent be delisted.  This patent has remained listed because, under Section 505(j)(5)(D)(i) of the Act, a first applicant may retain eligibility for 180-day exclusivity based on a paragraph IV certification to this patent for a certain period.  Applicants under Section 505(b)(2) are not required to certify to patents where this flag is set to Y.

```{r load_patents}
patents = read_csv("./data/patent.csv")
patents = patents %>% 
  clean_names() %>%
  rename(drug_type = appl_type) %>% 
  mutate(drug_type = recode(drug_type, "N" = "B")) %>%
  separate(patexp, into = c("month_pat_expires", "day_pat_expires", "year_pat_expires"), sep = "/") %>%
  separate(patent_no, into = c("patent_num", "pediatric_exclusivity_granted"), sep = "\\*")

patents$pediatric_exclusivity_granted[is.na(patents$pediatric_exclusivity_granted)] = "no"
patents$drugsub[is.na(patents$drugsub)] = "no"
patents$drugprod[is.na(patents$drugprod)] = "no"
patents$delist[is.na(patents$delist)] = "no"

patents = patents %>%
  mutate(pediatric_exclusivity_granted = factor(pediatric_exclusivity_granted, levels = c("no", "PED"), labels = c("No", "Yes"))) %>%
  mutate(drug_substance_flag = factor(drugsub, levels = c("no", "Y"), labels = c("no", "yes"))) %>%
  mutate(drug_prod_flag = factor(drugprod, levels = c("no", "Y"), labels = c("no", "yes"))) %>%
  mutate(patent_delisted = factor(delist, levels = c("no", "Y"), labels = c("no", "yes"))) %>%
  select(-drugsub, -drugprod, -patexpstr) %>%
  rename(pat_use_code = patuse)
patents
```

#### **Dataset 4: Products**

*The Products dataset from the FDA Orange Book

* Below is a description of the dataset and variables from https://www.fda.gov/Drugs/InformationOnDrugs/ucm129689.htm

* Ingredient: The active ingredient(s) for the product. Multiple ingredients are in alphabetical order, separated by a semicolon.

* Dose form;The product dosage form

* Route:  Product dosage route 

* Trade_Name: The trade name of the product as shown on the labeling.

* Applicant: The firm name holding legal responsibility for the new drug application.  The firm name is condensed to a maximum twenty character unique string.

* Strength: The potency of the active ingredient.  May repeat for multiple part products.

* Drug Application Type: The type of new drug application approval. NDA orANDA as for patents dataset.

* New Drug Application (NDA) Number: The FDA assigned number to the application.

* Product Number: The FDA assigned number to identify the application products. Each strength is a separate product.  May repeat for multiple part products.

* Therapeutic Equivalence (TE) Code: The TE Code indicates the therapeutic equivalence rating of generic to innovator Rx products.

* Approval Date: The date the product was approved as stated in the FDA approval letter to the applicant. Products approved prior to the January 1, 1982 contain the phrase: "Approved prior to Jan 1, 1982".  This separated into Year, Month, and Day variables.

* Reference Listed Drug (RLD) Updated: The RLD is a drug product approved under section 505(c) of the FD&C Act for which FDA has made a finding of safety and effectiveness. In the electronic Orange Book, an RLD is identified by “RLD” in the RLD column.

* Reference Standard (RS) New! :A “reference standard” is the drug product selected by FDA that an applicant seeking approval of an ANDA must use in conducting an in vivo bioequivalence study required for approval of an ANDA.  In the electronic Orange Book, a reference standard is identified by “RS” in the RS column.

*Type: The group or category of approved drugs. RX (prescription), OTC (over the counter), or DISCN (discontinued).

*Applicant Full Name: The full name of the firm holding legal responsibility for the new drug application. 

```{r load_products}
products = read_csv("./data/products.csv")
products = products %>%
  rename(drug_type = appl_type,
         otc = type) %>% 
  mutate(drug_type = recode(drug_type, "A" = "G", "N" = "B"),
         otc = recode(otc, "OTC" = "yes", "RX" = "no", "DISCN" = "no")) %>%
  separate(appdate, into = c("approval_month", "approval_day", "approval_year"), sep = "/") %>%
  #separate(doseform, into = c("form", "release_type"), sep = ",")
  mutate(otc = factor(otc)) %>%
  mutate(ref_listed_drug = factor(rld)) %>%
  mutate(ref_std = factor(rs)) %>%
  select(-appdatestr, -rld, -rs, -dfroute)
products
```

#### **Dataset 5: Exclusivity**

We load and tidy the `exclusivity` dataset

```{r tidy_exclusivity}
exclusivity = read_csv("./data/exclusivity.csv") %>%
  clean_names() %>%
  rename(drug_type = appl_type) %>% 
  mutate(drug_type = recode(drug_type, "A" = "G", "N" = "B")) %>% 
  separate(excldate, into = c("month_excl_expires", "date_excl_expires", "year_excl_expires"))
exclusivity
```

The tidied `exclusivity` dataset has `r exclusivity %>% nrow` observations and `r exclusivity %>% ncol` variables:

* `year`, `month`, and `date` the exclusivity expires
* `appl_type`: type of new drug application approval. New Drug Applications (NDA or innovator) are "N". Abbreviated New Drug Applications (ANDA or generic) are “A”.
* `appl_no`: FDA assigned number to the application.  Format is nnnnnn.
* `product_no`: FDA assigned number to identify the application products. Each strength is a separate product. May repeat for multiple part products.  Format is nnn.
* `exclcode`: Code to designate exclusivity granted by the FDA to a drug product.  Format is nnnnnnnnnn.
* `excldatestr`: The date the exclusivity expires. Format is MMM DD, YYYY.

#### **Dataset 6: Ingredient**

We load the `ingredient` dataset. 

```{r tidy_ingredients}
ingredient = read_csv("./data/ingredient.csv") %>%
  clean_names() %>%
  select(singnum, singredient, everything())
ingredient
```

The tidied `ingredient` dataset has `r ingredient %>% nrow` observations and `r ingredient %>% ncol` variables:

* `singnum`: The number assigned to the single ingredient for each product.
* `singredient`: each single ingredient in the product.
* `ingredient`: The active ingredient(s) for the product. Multiple ingredients are in alphabetical order, separated by a semicolon.

#### **Dataset 7: Drug by Condition**

Scraping data on drugs per condition from [this website](https://www.centerwatch.com/drug-information/fda-approved-drugs/medical-conditions/A)

```{r scrape_drug_disease}

urls = paste("https://www.centerwatch.com/drug-information/fda-approved-drugs/medical-conditions/", toupper(letters), sep = "")

get_drug_disease = function(url) {
  
  disease_data = read_html(url) %>% 
  html_nodes(css = ".ToggleDrugCategory") %>%
  html_text() %>% 
  as_tibble() %>% 
  mutate(key = value) %>% 
  rename(disease = value)
  
  drug_data = read_html(url) %>% 
  html_nodes(css = ".CategoryListSection a:nth-child(1)") %>%
  html_text() %>% 
  as_tibble() %>% 
  mutate(key = value) %>% 
  rename(drug = value)
  
  drug_disease_data = read_html(url) %>% 
  html_nodes(css = "#MainColumn a:nth-child(1)") %>%
  html_text() %>% 
  as_tibble() %>% 
  mutate(key = value) %>% 
  rename(drug_disease = value)
  
  drug_disease_joined = left_join(drug_disease_data, drug_data, disease_data, by = "key") %>% 
  left_join(., disease_data, by = "key") %>%
  fill(disease) %>%
  select(-c(drug,drug_disease))
  
  toremove = drug_disease_joined %>% distinct(disease) %>% pull()
  
  drug_disease_joined = drug_disease_joined %>% 
  filter(!key %in% toremove) %>% 
  rename(drug = key) %>% 
  .[-c(1:29),]
  
  drug_disease_joined
}

drug_disease_data = map_df(urls, get_drug_disease) %>% 
  mutate(disease = str_to_title(disease),
         drug = str_to_title(drug)) %>% 
  rename(trade_name = drug) %>%
  mutate(trade_name = trimws(trade_name, which = c("both"))) %>% 
  separate(trade_name, into = c("trade_name", "remove"), sep = " ") %>% 
  select(-remove)

drug_disease_data
```

The `drug_disease_data` contains `r drug_disease_data %>% nrow` observations and `r drug_disease_data %>%  ncol` variables:

* `drug`: the drug name
* `disease`: the disease name


#### **Dataset 8: Chronic Diseases**

```{r}
#first dataset
chronic_diseases_1_html = read_html("https://www.medicalschemes.com/medical_schemes_pmb/chronic_disease_list.htm")
chronic_diseases_1 = chronic_diseases_1_html %>%
  html_nodes("td td td .maintext") %>%
  html_text() %>%
  as_tibble() %>%
  rename(disease = value)

#second dataset
chronic_diseases_2_html = read_html("http://www.health24.com/Medical-schemes/PMB-and-chronic-disease/List-of-chronic-diseases-20120721")
chronic_diseases_2 = chronic_diseases_2_html %>%
  html_nodes("ul:nth-child(10) li , ul:nth-child(10) a") %>%
  html_text() %>%
  as_tibble() %>%
  rename(disease = value)

#third dataset
chronic_diseases_3_html = read_html("https://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/Chronic-Conditions/CC_Main.html")
chronic_diseases_3 = chronic_diseases_3_html %>%
  html_nodes("td") %>%
  html_text() %>%
  as_tibble() %>%
  rename(disease = value)

#final chronic disease dataset
chronic_diseases = rbind(chronic_diseases_1, chronic_diseases_2, chronic_diseases_3) %>%
  distinct()

#recode diseases
chronic_diseases$disease[62] = "Cancer"
chronic_diseases$disease[67] = "High Blood Pressure"
chronic_diseases$disease[43] = "High cholesterol"
chronic_diseases$disease[60] = "Austism"
chronic_diseases$disease[55] = "Arthritis"
chronic_diseases$disease[46] = "Hypothyroidism"
chronic_diseases$disease[38] = "Dysrhythmia"
chronic_diseases$disease[66] = "inactive thyroid gland"
chronic_diseases$disease[53] = "alzheimer's disease"
chronic_diseases$disease[26] = "Dementia"
chronic_diseases$disease[56] = "Hepatitis"
chronic_diseases$disease[71] = "COPD"
```

#### Join Datasets

```{r join_dataset}
#A: join datasets by ingredient for ingredient and products (left join)
#products = longer dataset + ingredient
product_ingredient = left_join(products, ingredient, by = "ingredient")
product_ingredient

#B: exclusivity and patents left join (more~less)
patent_exclusivity = left_join(patents, exclusivity, by = c("appl_no", "product_no", "drug_type"))
patent_exclusivity

#C: join A and B left join: "Orange Book." Thursday: meet to eliminate variables (inner join: only commonalities are preserved) #eliminate drugs that do not match
orange_book = left_join(product_ingredient, patent_exclusivity, by = c("appl_no", "product_no", "drug_type"))

orange_book = orange_book %>%
  mutate(ingredient = str_to_title(ingredient),
         doseform = str_to_title(doseform),
         route = str_to_title(route),
         trade_name = str_to_title(trade_name),
         applicant = str_to_title(applicant),
         otc = str_to_title(otc),
         appfull = str_to_title(appfull),
         singredient = str_to_title(singredient),
         strength = str_to_title(strength))
orange_book

# final goal: disease, funding, drug prices, and Orange Book

# eliminate unnecessary variables from orange book dataset
orange_book_clean = orange_book %>%
  rename(application_no = appl_no, 
         active_ingredient = singredient, 
         dose_form = doseform) %>%
  select(-c(ingredient, applicant, product_no, tecodestr, tecode, approval_day, ref_std, singnum, day_pat_expires, pat_use_code, drug_substance_flag, drug_prod_flag, exclcode, excldatestr, date_excl_expires)) %>% 
  select(trade_name, everything()) %>% 
  separate(dose_form, into = c("dose_form", "remove")) %>%
  filter(is.na(remove)) %>% 
  select(-remove) %>% 
  filter(str_detect(dose_form, regex("Tablet", ignore_case = TRUE))) %>%
  mutate(trade_name = trimws(trade_name, which = c("both")),
         trade_name = str_replace_all(trade_name, ", ", "-"),
         trade_name = str_replace_all(trade_name, " And ", "-"))

# inner join of disease drug and orange book by trade_name
final_dataset =
  new_drug_price %>%
  inner_join(., orange_book_clean, by = c("trade_name", "drug_type", "otc")) %>% 
  distinct() %>% 
  left_join(., drug_disease_data, by = "trade_name") %>% 
  distinct() %>% 
  mutate(disease = str_replace(disease, " And ", "_"),
         disease = str_replace(disease, "High Blood Pressure \\(", ""),
         disease = str_replace(disease, "Copd \\(", ""),
         disease = str_replace(disease, "\\/Hyperactivity", ""),
         disease = str_replace(disease, "\\)", ""),
         disease = str_replace(disease, " Mellitus, Type 2", ""),
         disease = str_replace(disease, " Mellitus Types I", ""),
         disease = str_replace(disease, "Hepatitis", "Hepatitis C"),
         disease = str_replace(disease, "Hepatitis C B", "Hepatitis B"),
         disease = str_replace(disease, "Hepatitis C C", "Hepatitis C"),
         disease = str_replace(disease, "Hiv/Aids", "HIV/AIDS"),
         disease = str_replace(disease, " \\- ", "_"),
         disease = str_replace(disease, "Migraine", "Migraines")
         ) %>%
  separate(disease, into = c("disease", "remove"), sep = "\\_") %>% 
  select(-remove) %>% 
  separate(disease, into = c("disease", "remove"), sep = "\\(") %>% 
  select(-remove) %>% 
  separate(disease, into = c("disease", "remove"), sep = "\\;") %>%
  select(-remove) %>% 
  left_join(., final_NIH_funding, by = c("disease", "year")) %>%
  as_tibble() %>%
  distinct() %>% 
  mutate(disease = str_replace(disease, "Chronic Obstructive Pulmonary Disease", "Chronic Obstructive Pulmonary Disorder"),
         disease = str_replace(disease, "Kidney Disease", "Chronic Kidney Disease"),
         disease = str_replace(disease, "Cardiac Ischemia", "Coronary Artery Disease"),
         disease = str_replace(disease, "Diabetes", "Diabetes Mellitus Types 1 & 2"),
         disease = str_replace(disease, "Bipolar Disorder", "Bipolar Mood Disorder"))

#joining for chronic
#need to include breast cancer, hepatitis b, hepatitis c, lung cancer, prostate cancer bc chronic disease set does not specify
#hepatitis
hepatitis_chronic = final_dataset %>%
  filter(str_detect(disease, regex("hepatitis", ignore_case = TRUE))) %>%
  pull(disease) %>%
  unique() %>%
  as_tibble() %>%
  rename(disease = value)

#cancer
cancer_chronic = final_dataset %>%
  filter(str_detect(disease, regex("cancer", ignore_case = TRUE))) %>%
  pull(disease) %>%
  unique() %>%
  as_tibble() %>%
  rename(disease = value)

#hiv/aids
hiv_aids_chronic = final_dataset %>%
  filter(str_detect(disease, regex("hiv", ignore_case = TRUE)) | str_detect(disease, regex("aids", ignore_case = TRUE))) %>%
  pull(disease) %>%
  unique() %>%
  as_tibble() %>%
  rename(disease = value)

#combine hepatitis, cancer and hiv/aids with chronic diseases
chronic_diseases = rbind(chronic_diseases, hepatitis_chronic, cancer_chronic, hiv_aids_chronic)
chronic_diseases = chronic_diseases %>%
  unique()

#indicator of chronic disease
indicator_chronic_disease = vector()
indicator_chronic_disease[1:nrow(chronic_diseases)] = c("yes")
indicator_chronic_disease = as.tibble(indicator_chronic_disease)

#final chronic disease dataset with indicator column (y/n)
chronic_diseases_final = cbind(chronic_diseases, indicator_chronic_disease) %>%
  rename(chronic_disease = value) 
#formatting for exact matching
final_dataset$disease = trimws(final_dataset$disease, which = "right")
chronic_diseases_final$disease = trimws(chronic_diseases_final$disease, which = "both")
final_dataset$disease = tolower(final_dataset$disease)
chronic_diseases_final$disease = tolower(chronic_diseases_final$disease)

# Join final dataset and chronic disease dataset to complete analysis.
final_dataset = left_join(final_dataset,chronic_diseases_final, by = "disease")
final_dataset$chronic_disease[is.na(final_dataset$chronic_disease)] = "no"
final_dataset = final_dataset %>%
  mutate(disease_type = as.factor(recode(chronic_disease, 'no' = "acute", 'yes' = "chronic"))) 

rm(chronic_diseases, indicator_chronic_disease, cancer_chronic, hepatitis_chronic, hiv_aids_chronic, chronic_diseases_1, chronic_diseases_2, chronic_diseases_3, chronic_diseases_1_html, chronic_diseases_2_html, chronic_diseases_3_html, chronic_diseases_final, patent_exclusivity, product_ingredient) #clean up workspace
```

### **Case Study** : description and summary statistics of most advertised vs. most abused vs. most prescribed 

#### Most Advertised

#### Most Abused

#### Most Prescribed



### Analysis & Discussion

#### **Josie**
```{r}
top_advertising_cost = c("Viagra")

#shiny application : most advertised (by spending) vs. most abused vs. most prescribed (by sales)
#another shiny application: give list of all "most" drugs and can see boxplot for that specific drug
final_dataset %>%
  filter(!is.na(disease) & !is.na(avg_price)) %>%
  filter(trade_name %in% top_advertising_cost) %>%
  group_by(trade_name) %>% 
  plot_ly(y = ~avg_price, type = "box", color = ~trade_name, colors = "Set3") %>%
  layout(title = "Distribution of Average Price for Top Advertised Drugs")

#shiny application: acute vs. chronic
final_dataset %>%
  filter(!is.na(disease) & !is.na(funding)) %>%
  mutate(disease = factor(disease)) %>%
  group_by(disease) %>%
  summarize(mean_funding = mean(funding)) %>%
  mutate(disease = fct_reorder(disease, mean_funding)) %>%
  plot_ly(x = ~disease, y = ~mean_funding, type = "bar")
```

#### **Manali**

Next, we looked to explore whether there was a relationship between prevalence and funding.  While the number of diseases with the required complete data was relatively small, it appears from the scatterplot that when you have a low prevalence, there is a wide range of the average funding.  As the prevalence increases the funding amount generally seems to be on the lower side.  However, if you look more closely at which diseases are included, some interesting observations can be made. Many of the diseases with high funding are cancers, which have a low prevalence.  This makes sense as there is a lot of research focused on cancer and new drug development.  Most cancers have large subgroups with no good treatment options, so it makes sense that a large amount of funding is devoted to cancer.  Some diseases with higher prevalence but lower funding are hypertension, migraines, ADD, and COPD, which do already have treatment options available on the market.

Then, we looked to determine whether funding correlated with drug prices.  This does not really seem to be the case from the scatterplot.  Along the entire range of funding amounts, most drugs still have relatively low prices.  It is interesting that different drugs that treat the same disease seem to be largely clustered together.  We see breast cancer treatments are unusually high in price, which again makes sense given what we know about cancer treatment. 
```{r}
#does funding correlate with drug prices?

final_dataset %>%
  #sample_n(50)%>%
  group_by(disease, funding, avg_price) %>%
  summarize(avg_funding = mean(funding, na.rm = TRUE)) %>%
  #plot_ly(x = ~funding, y = ~avg_price, type = "scatter")
  plot_ly(x = ~ avg_funding, y = ~ avg_price, type = "scatter", color = ~disease, text = ~paste('Disease:', disease))
```


#### **Muhire**

* We look at the yearly distribution of drug prices for a few drugs. Some drugs have skewed price distributions and the mean (instead of the mean) price might be the better measure to use in the analyses. Also, it might be worth creating a shiny app for people to view the price distributions for various drugs.

```{r drug_price_distr}
drug_price %>%
  filter(trade_name == "Abacavir") %>%
  ggplot(aes(x = year, y = unit_price)) + geom_violin(aes(fill = year), color = "blue", alpha = .5)
```

We created a spaghetti plot of the average price of drugs per year for each drug. We only used data on prices per drug unit (mainly per tablet). Information about the disease that the drugs treat can be seen by hovering over the plot. We notice that of the drugs with information on price per unit drug, the drugs for Hepatitis C seem to have the highest cost.

**Possible case studies on most popular (perceived) drugs: *Lipitor (Schizophrenia; not in dataset)*, Xanax (Panic Disorders, not in dataset), Celebrex (arthritis; not in dataset), Prozac (Depression; present in data. Price increased), Crestor, Nexium, Lisinopril, Zoloft (Panic Disorders; present in dataset. Almost doubled from 2013 to 2017)**

```{r price_drug_disease}
final_dataset %>% 
  filter(disease %in% c("Obesity")) %>%
  filter(pricing_unit == "EA" & !is.na(disease)) %>% 
  plot_ly(x = ~year, y = ~avg_price, type = "scatter", mode = "lines", text = ~paste("Drug: ", trade_name, '<br>Disease:', disease), color = ~trade_name, alpha = 0.5) %>% 
  layout(showlegend = FALSE)
```

We also created a scatter plot of the mean NIH funding per drug (assumed to be the same as the NIH funding for disease research) against the demand of the drug (defined as the 2015 disease prevalence) colored by disease. Among drugs with available information on price per unit drug, obesity seems to have received the most funding on average and to have the highest 2015 prevalence.

```{r funding_demand}
final_dataset %>% 
  filter(pricing_unit == "EA" & !is.na(funding) & !is.na(prevalence)) %>% 
  group_by(trade_name, disease, prevalence) %>% 
  summarize(mean_fund = mean(funding)) %>% 
  plot_ly(x = ~prevalence, y = ~mean_fund, type = "scatter", alpha = 1, text = ~paste("Drug: ", trade_name, '<br>Disease:', disease), color = ~disease) %>%
  layout(showlegend = FALSE)
```

#### **MeOak**
```{r}

```

### Project Website
```{r project_website}
#project website https://josiealford14.github.io/therealcostofdrugs/
```

### Conclusion

