---
title: "Patents of Pharmaceutical Companies"
output: 
  html_document:
    code_folding: hide
---
#
#
```{r load_packages, echo = FALSE, results = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidytext)

library(broom)
library(dplyr)
library(forcats)
library(httr)
library(janitor)
library(knitr)
library(magrittr)
library(plotly)
library(readxl)
library(rvest)
library(stringr)
library(viridis)
```

```{r load_data, echo = FALSE, results = FALSE, warning = FALSE, message = FALSE}
final_dataset = read_csv("./data/final_dataset_20171205.csv") 
final_NIH_funding = read_csv("./data/final_NIH_funding.csv") 
drug_price = read_csv("../data/drug_price.csv") 
#Patent Cleaning
patents = final_dataset %>%
  ungroup() %>% 
  select(appfull, disease, avg_price, patent_num, year_pat_expires, approval_year, year_excl_expires) %>% 
  filter(!duplicated(patent_num)) %>%
  filter(!is.na(disease)) %>%
  filter(year_pat_expires > 2017) %>% 
  mutate(
    year_pat_expires = as.factor(year_pat_expires), 
    disease = str_replace(disease, "Clostridium Difficile-Associated Diarrhea", "Diarrhea_nosocomial"),
    disease = str_replace(disease, "Chronic Diarrhea", "Diarrhea (chronic)"),
    disease = str_replace(disease, "Chronic Kidney Disease", "Kidney Disease (chronic)"),
    disease = str_replace(disease, "Chronic Myeloid Leukemia", "Myeloid Leukemia"),
    disease = str_replace(disease, "Thyroid Disease", "Thyroid Disorders"),
    #appfull = str_replace(appfull, "-", " "),
    appfull = str_replace(appfull, " Ab", ""),
    appfull = str_replace(appfull, " And", ""),
    appfull = str_replace(appfull, " Biotech", ""),
    appfull = str_replace(appfull, " Co", ""),
    appfull = str_replace(appfull, " Inc", ""),
    appfull = str_replace(appfull, " International", ""),
    appfull = str_replace(appfull, " Ireland s", ""),
    appfull = str_replace(appfull, " Ltd", ""),
    appfull = str_replace(appfull, " Llc", ""),
    appfull = str_replace(appfull, " Lp", ""),
    appfull = str_replace(appfull, " Holdings", ""),
    appfull = str_replace(appfull, " Pharmaceutical", ""),
    appfull = str_replace(appfull, " Pharmaceuticals", ""),
    appfull = str_replace(appfull, " Products", ""),
    appfull = str_replace(appfull, " Research Institute", ""),
    appfull = str_replace(appfull, " Research Development", ""),
    appfull = str_replace(appfull, " s ", ""),
    appfull = str_replace(appfull, " Trading", ""),
    appfull = str_replace(appfull, " Unltd", ""),
    appfull = str_replace(appfull, "Allergans", "Allergan"),
    appfull = str_replace(appfull, "Astrazenecas", "Astrazeneca"),
    appfull = str_replace(appfull, "Bristol-Myers Squibb", "Bristol Myers Squibb"),
    appfull = str_replace(appfull, "Bristol-Myers Squibb Co", "Bristol Myers Squibb Co"),
    appfull = str_replace(appfull, "Janssens", "Janssen"),
    appfull = str_replace(appfull, "Ingelheims", "Ingelheim"),
    appfull = str_replace(appfull, "Pfizers", "Pfizer"),
    appfull = str_replace(appfull, "Wyeths Wholly Owned Sub Pfizer Inc", "Pfizer")) %>%
  arrange(appfull)

#Labeling Axes
xax_pharma = list(
  title = "Pharmaceutical Company",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)
xax_disease = list(
  title = "Disease",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = FALSE
)
yax = list(
  title = "No. of Patents",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = TRUE,
  showgrid = TRUE
)
yax_1 = list(
  title = "No. of Patents",
  zeroline = FALSE,
  showline = FALSE,
  showticklabels = FALSE,
  showgrid = TRUE
)
```

### The number of patents currently held by pharmaceutical companies 
```{r, warning = FALSE, message = FALSE}
#Avg Drug Prices by Company for each Disease
pharma_charge = patents %>% 
  group_by(disease, appfull) %>%
  summarize(comp_avg_price = mean(avg_price)) 

pharma_charge %>% 
  plot_ly(x = ~comp_avg_price, y = ~disease, type = "scatter",  text = ~paste('Company: ', appfull, '<br> Disease', disease), color = "red") %>% 
  layout(title = "Avg Drug Price By Disease", 
         yaxis = list(
           title = "",
           zeroline = FALSE,
           showline = FALSE,
           showticklabels = TRUE,
           showgrid = TRUE))
```

### Current Patents and Projection of expiring patents by company
```{r, warning = FALSE, message = FALSE}
#No. of Patents Currently Held by Pharma Company
patents %>%
  group_by(appfull) %>%
  count() %>%
  rename(num_pat_held = n) %>%
  ungroup() %>%
  mutate(appfull = fct_reorder(appfull, num_pat_held)) %>%
  plot_ly(x = ~appfull, y = ~num_pat_held, color = ~appfull, type = "bar", colors = "blue") %>% 
  layout(title = "No. of Patents Currently Held by Pharmaceutical Company",  yaxis = yax, showlegend = FALSE)
#xaxis = xax_pharma,
```
AstraZeneca is currently the largest patent holder.
Keryx Pharmaceuticals has the most number of patents expiring overall.  Keryx's 11 patents expiring in 2024 cover chronic kidney disease allowing generics to enter the kidney disease markt.


### The number of patents currently held by each specfic disease
```{r, warning = FALSE, message = FALSE}
#Projection of Expiring Patents by Pharma Company
pat_exp = patents %>%
  group_by(appfull, year_pat_expires) %>%
  count() %>% 
  rename(num_pat_exp = n)
#No. of Patents Currently Held by Disease
patents %>%
  group_by(disease) %>%
  count() %>%
  rename(num_pat_held = n) %>%
  ungroup() %>%
  mutate(disease = fct_reorder(disease, num_pat_held)) %>%
  plot_ly(x = ~disease, y = ~num_pat_held, color = ~disease, type = "bar", colors = "red") %>% 
  layout(title = "No. of Patents Currently Held by Disease",  yaxis = yax, showlegend = FALSE)
#xaxis = xax_disease,
```
### Projection of expiring patents per condition or disease
```{r}
#Projection of Expiring Patents by Disease
patents %>%
  group_by(disease, year_pat_expires) %>%
  count() %>%
  rename(num_pat_exp = n) %>%
  # ungroup() %>% 
  plot_ly(x = ~year_pat_expires, 
          y = ~disease, 
          color = ~disease, 
          type = "scatter",
          colors = "Set3", 
          mode = 'markers', 
          marker = list(size = ~num_pat_exp*3), 
          text = ~paste('Disease:', disease, 'Count:', num_pat_exp)) %>%
  layout(title = "Projection of Expiring Patents by Disease", 
           yaxis = list(
           title = "",
           zeroline = FALSE,
           showline = FALSE,
           showticklabels = TRUE,
           showgrid = TRUE), 
         showlegend = FALSE)
```
The disease with the most patents expiring over the next several years chronic kidney disease.